# Smart-Data-Predictor Exploratory Data Analysis

**This Exploratory Data Analysis (EDA) is an essential step in the model-building process, focusing on identifying patterns in the data-set and examining the key factors that influence the price of a used car.**

## **Objective**

The goal of this EDA is to analyze trends and insights within the used car dataset to understand the factors affecting car prices. Through **data preprocessing, visualization, and statistical analysis**, we will explore relationships between key variables such as:

-   **Year of manufacture**

-   **Kilometers driven**

-   **Fuel type**

-   **Transmission type**

-   **Ownership history**

The insights from this analysis will guide **feature selection** and enhance the development of a **robust price prediction model**.

## Libraries Used

```{r}
library(dplyr) # data manipulation
library(ggplot2) # data visualization
library(tidyr) # data tidying and reshaping
library(scales) # scale balancing
library(ggcorplot)
```

```{r}
#VISUALIZATION SUPPORT

```

## Data Collection

```{r}
used_cars <- readr::read_csv("all_cars.csv") 
```

## Data Information

### Data Size & Structure

```{r}
dim(all_cars)
str(all_cars)

```

## Data Insights to generate

These are the following insights we are generate

1.  Distribution of numbers of car owners

2.  Average selling price by car owners

3.  Distribution of Fuel type

4.  Average selling price by fuel type

5.  Distribution of Price.

6.  Relationship between kilometer driven and selling price

```{r}
lapply(colnames(all_cars), is.character)
```

```{r}
# 1. Distribution of Car Owners
# Group by ownership type and count occurrences
owner_distribution <- all_cars %>%
  group_by(no_own) %>%
  summarise(owner_count = n())  %>%  arrange(desc(owner_count))

# color
colors <- colorRampPalette(c("#FFFF00", "#FF0000"))(nrow(owner_distribution))

owner_distribution
# Create a bar plot
barplot(owner_distribution$owner_count, 
        names.arg = owner_distribution$no_own, 
        col = colors,
        main = "Distribution of Number of Previous Owners",
        xlab = "Number of Owners",
        ylab = "Count")
abline(h = mean(owner_distribution$owner_count), lty = 2, lwd = 2)


```

```{r}
# 2. Average selling price by car owners 
owner_price <- all_cars %>%
  group_by(no_own) %>%
  summarise(avg_price = mean(price)) %>%
  arrange(desc(avg_price))

# Define color gradient with valid hex codes
colors_blyl <- colorRampPalette(c("#0000FF", "#ADD8E6"))(nrow(owner_price))

# Plotting the barplot
barplot(owner_price$avg_price, 
        names.arg = owner_price$no_own, 
        col = colors_blyl,
        main = "Distribution of Price by Previous Owners",
        xlab = "Number of Owners",
        ylab = "Price",
     ylim = c(0, max(owner_price$avg_price) * 1.1))  # Use avg_price for y-axis limit
 

# Adding a horizontal line at the average price across all data
abline(h = mean(owner_price$avg_price), lty = 2, lwd = 2)


```

```{r}

# 3. Group by fuel type and count occurrences
fuel_distribution <- all_cars %>%
  group_by(fuel_type) %>%
  summarise(fuel_count = n()) %>%
  arrange(desc(fuel_count))

# Define a color gradient
colors <- colorRampPalette(c("#FF5733", "#FFC300", "#DAF7A6"))(nrow(fuel_distribution))

# Create a bar plot with colors
barplot(fuel_distribution$fuel_count, 
        names.arg = fuel_distribution$fuel_type,
        col = colors,  # Apply the color gradient
        main = "Distribution of Fuel Type",
        xlab = "Fuel Type",
        ylab = "Count")




```

```{r}
# 4. Average selling price by fuel type 
fuel_price <- all_cars %>%
  group_by(fuel_type) %>%
  summarise(avg_price = round(mean(price), 0)) %>%
  arrange(desc(avg_price))

# Define color gradient
colors_blyl <- colorRampPalette(c("#0000FF", "#ADD8E6"))(nrow(fuel_price))

# Plotting the barplot
barplot(fuel_price$avg_price, 
        names.arg = fuel_price$fuel_type, 
        col = colors_blyl,
        main = "Average Selling Price by Fuel Type",
        xlab = "Fuel Type",
        ylab = "Average Price",
        ylim = c(0, max(fuel_price$avg_price) * 1.4))  # Use avg_price for y-axis limit

# Adding a horizontal line at the overall average price
abline(h = mean(fuel_price$avg_price), lty = 2, lwd = 2)

options(scipen = 999)
  
ggplot(fuel_price, aes(x = reorder(fuel_type, -avg_price), y = avg_price)) + 
  geom_bar(stat = "identity", fill = "lightblue", color = "black") +
  labs(title = "Average Selling Price by Fuel Type", 
       x = "Fuel Type", 
       y = "Average Price") +
  theme_minimal()



```

## Price Distribution

```{r}
# Distribution of Price
ggplot(all_cars, aes(x = price)) + 
  geom_histogram(bins = 45, fill = "lightblue", color = "black") +
  geom_vline(xintercept = mean(all_cars$price), 
             linetype = "dashed", color = "red", lwd = 1) +
    geom_vline(xintercept = median(all_cars$price), 
             linetype = "dashed", color = "blue ", lty = 3, lwd = 1) +
  labs(title = "Distribution of Car Prices", x = "Price", y = "Count") +
  theme_minimal()

```

## Relationship Between Kilometer Driven and Price

```{r}
ggplot(all_cars, aes(x = Year_mfd, y = price, color = fuel_type)) +
  geom_point(alpha = 1.5) +
  labs(title = "Price Vs Year Manufactured",
       x = "Year Manufactured",
       y = "Price") +
  theme_bw()
```

## Correlation Between Price and Kilometer Driven

```{r}
# Scatter plot to show the relationship between price and kilometers driven
ggplot(all_cars, aes(x = km_drv, y = price)) + 
  geom_point(alpha = 0.5, color = "blue") +  # Points with some transparency
  labs(title = "Correlation Between Price and Kilometer Driven",
       x = "Kilometers Driven",
       y = "Price") +
  theme_minimal()
D365TDH
```

```{r}
cor_matrix <- cor(all_cars[, c("Year_mfd", "km_drv", "price")]) #Select numerical cols

ggcorplot(cor_matrix, method = "square", digits = 2, lab = TRUE)
```
